Quantum Walks and Application to Quantum Money\\[1em] \normalsize McMaster University\\ Department of Computing and Software

### 2025-04-13 draft.tex
draft updated

@@

Interestingly, there exist formal \textit{equivalences} between the two models. A coined DTQW can approximate a CTQW through small time-step evolution, as demonstrated by Childs et al., and conversely, CTQWs can often be embedded into DTQWs by carefully engineering the coin space to emulate continuous dynamics. While both are computationally equivalent in power, each model has practical advantages in different settings. CTQWs are frequently preferred for spatial search on low-degree graphs due to the natural formulation of marked Hamiltonians. DTQWs, especially in Szegedy’s coinless formulation, excel in quantizing classical Markov chains and have been instrumental in algorithms such as element distinctness.

@@

% \[
% A = \sum_{s \in S} P_s,
% \]
% where $P_s$ is the permutation matrix representing left multiplication by $s \in G$: $P_s \ket{g} = \ket{sg}$ for all $g \in G$.

% If $S$ is symmetric (i.e., closed under inversion), then $A$ is Hermitian, and $P_s^\dagger = P_{s^{-1}}$. The eigenstates of $A$ are understood through group representation theory. In particular, the regular representation of $G$ on $\mathbb{C}^{|G|}$ decomposes into a direct sum of irreducible representations (irreps), and since each $P_s$ acts via the regular representation, the adjacency matrix $A$ is a sum of representation operators:
% \[
% A = \sum_{s \in S} \rho(s),
% \]
% where $\rho$ denotes the regular representation.

% By Schur’s lemma, $A$ acts within each irreducible subspace as a matrix proportional to the representation of $S$. Hence, $A$ is block-diagonal (in a suitable basis), with blocks corresponding to irreps of $G$. If $\{\phi_\alpha\}$ are the irreps of $G$, then the eigenvalues of $A$ can be determined from the character values $\chi_\alpha(s)$:
% \[
% \lambda_\alpha = \sum_{s \in S} \chi_\alpha(s), \quad \text{(for abelian $G$, $\chi_\alpha$ is a character)}.
% \]

% In the case of abelian $G$, all irreps are one-dimensional, and the eigenstates of $A$ are the Fourier basis:
% \[
% \ket{\chi} = \frac{1}{\sqrt{|G|}} \sum_{g \in G} \chi(g) \ket{g}, \quad \lambda_\chi = \sum_{s \in S} \chi(s).
% \]

% \textbf{Example – Hypercube:} For $G = \mathbb{Z}_2^n$, characters are labeled by bitstrings $u \in \{0,1\}^n$, and $\chi_u(g) = (-1)^{u \cdot g}$. Then
% \[
% \lambda_u = \sum_{j=1}^n (-1)^{u_j},
% \]
% which gives the $n+1$ distinct eigenvalues $n, n-2, \dots, -n$ of the $n$-dimensional hypercube.

% In the nonabelian case, similar principles apply but require using higher-dimensional irreps and the Great Orthogonality Theorem.

@@

% \paragraph{Hitting and Mixing Properties}

% Quantum walks on Cayley graphs have been applied to study hitting and mixing times. In the classical case, random walks on many Cayley graphs are rapidly mixing (e.g., those based on nonabelian simple groups, which often form expanders).

% In the quantum case, mixing is more subtle due to unitary evolution. One typically considers \emph{instantaneous} or \emph{time-averaged} mixing. Some Cayley graphs support perfect state transfer or periodicity in CTQWs.

% \textbf{Examples:}
% \begin{itemize}
%     \item On the cycle graph $C_n$, a CTQW exhibits perfect revival at time $t = \pi$ if $n$ is even.
%     \item On the hypercube, the CTQW exhibits approximate uniform mixing at specific times.
% \end{itemize}

% These properties are derived from the spectral structure, which is accessible via group characters.

% \paragraph{Discrete-Time Walks and Symmetry}

% Coined DTQWs also benefit from Cayley graph symmetry. If a symmetric coin (e.g., the Grover coin) is used at each vertex of a vertex-transitive graph, the walk respects the full symmetry. As a result:
% \begin{itemize}
%     \item Amplitudes often remain uniform on symmetry-equivalent vertices.
%     \item The dynamics of the walk can often be simplified by restricting to a smaller symmetric subspace.
%     \item In some cases, symmetry can lead to \emph{localization}, where amplitude becomes trapped due to spectral properties of the coin.
% \end{itemize}

% These effects can be mitigated or exploited through careful coin design or by using Szegedy's method, which naturally respects symmetry.

% In summary, Cayley graphs provide a rich and algebraically structured setting for quantum walks. Their underlying group-theoretic properties enable powerful techniques such as spectral analysis through irreducible representations, dimensionality reduction via symmetries and invariant subspaces, and explicit calculations of probabilities, mixing behavior, and hitting times. These features make Cayley graphs especially amenable to efficient quantum walk algorithms, which often exploit such structures either explicitly or implicitly. In the sections that follow, we extend these ideas to more general settings where group actions go beyond the framework of simple Cayley graphs.

\paragraph{Hitting and Mixing Properties}
Quantum walks on Cayley graphs have been applied to study hitting and mixing times. In the classical case, random walks on many Cayley graphs are rapidly mixing (e.g., those based on nonabelian simple groups, which often form expanders).

In the quantum case, mixing is more subtle due to unitary evolution. One typically considers instantaneous or time-averaged mixing. Some Cayley graphs support perfect state transfer or periodicity in CTQWs~\cite{Gerhardt2003}. Recent work has also established upper bounds on the mixing time for quantum walks on random graphs, showing polynomial-time convergence in certain regimes~\cite{Chakraborty2020}.

Examples:
\begin{itemize}
    \item On the cycle graph $C_n$, a CTQW exhibits perfect revival at time $t = \pi$ if $n$ is even.
    \item On the hypercube, the CTQW exhibits approximate uniform mixing at specific times.
\end{itemize}
These properties are derived from the spectral structure, which is accessible via group characters.

@@

To see this, we have:

>>>>

% \section{Mixing in Classical and Quantum Walks}

% \paragraph{Classical Mixing Time}

% The \emph{mixing time} of a random walk on a graph quantifies how quickly the probability distribution over vertices converges to the stationary distribution. Consider a continuous-time random walk on a graph \( G = (V, E) \) defined via the Laplacian matrix \( L \). The evolution of the probability vector \( p(t) \in \mathbb{R}^{|V|} \), which represents the probability of being at each vertex at time \( t \), is governed by the differential equation:
% \[
% \frac{dp(t)}{dt} = Lp(t).
% \]
% Let \( u := \frac{1}{|V|}(1,1,\dots,1) \) denote the uniform distribution over vertices. This vector is an eigenvector of \( L \) corresponding to eigenvalue zero. Assuming \( G \) is connected, this eigenvalue is unique. Let \( \{ v_\lambda \} \) be an orthonormal basis of eigenvectors of \( L \), where \( \lambda \neq 0 \), such that:
% \[
% L = \sum_{\lambda \neq 0} \lambda v_\lambda v_\lambda^\top.
% \]
% Then, the solution to the walk is:
% \begin{align}
% p(t) &= e^{Lt}p(0) \label{eq:classical1} \\
%      &= \left( |V| uu^\top + \sum_{\lambda \neq 0} e^{\lambda t} v_\lambda v_\lambda^\top \right)p(0) \label{eq:classical2} \\
%      &= \langle u, p(0) \rangle u + \sum_{\lambda \neq 0} e^{\lambda t} \langle v_\lambda, p(0) \rangle v_\lambda. \label{eq:classical3}
% \end{align}
% As \( t \to \infty \), each term involving \( \lambda < 0 \) decays exponentially, and \( p(t) \to u \). Hence, the walk asymptotically approaches the uniform distribution. The rate of convergence is governed by the \emph{spectral gap}, i.e., the magnitude of the smallest nonzero eigenvalue of \( L \).

% \paragraph{Quantum Mixing: Time-Averaged Behavior}

% In contrast to classical walks, a quantum walk is a unitary process. As such, it does not converge to a stationary quantum state over time. However, it is still meaningful to define a limiting distribution in the measurement statistics. Consider a quantum walk generated by a Hermitian operator \( H \) (e.g., the Laplacian or adjacency matrix), evolving under:
% \[
% \ket{\psi(t)} = e^{-iHt} \ket{a},
% \]
% where \( \ket{a} \) is the initial vertex state. Since unitary evolution is reversible and norm-preserving, convergence is not possible in the traditional sense. Nevertheless, one may define a time-averaged probability of observing vertex \( b \) after evolving for a randomly chosen time \( t \in [0, T] \):
% \[
% p_{a \to b}(T) = \frac{1}{T} \int_0^T |\langle b | e^{-iHt} | a \rangle|^2 \, dt.
% \]
% Assuming a non-degenerate spectral decomposition \( H = \sum_\lambda \lambda \ket{\lambda}\bra{\lambda} \), one finds:
% \begin{align}
% p_{a \to b}(T) &= \sum_\lambda |\langle a | \lambda \rangle|^2 |\langle b | \lambda \rangle|^2 \\
% &\quad + \sum_{\lambda \neq \lambda'} \langle b | \lambda \rangle \langle \lambda | a \rangle \langle a | \lambda' \rangle \langle \lambda' | b \rangle \cdot \frac{1 - e^{-i(\lambda - \lambda')T}}{i(\lambda - \lambda')T}.
% \end{align}
% As \( T \to \infty \), the second term vanishes, and the limiting distribution becomes:
% \[
% p_{a \to b}(\infty) := \sum_\lambda |\langle a | \lambda \rangle|^2 |\langle b | \lambda \rangle|^2.
% \]
% This distribution does not depend on time and captures the long-time average behavior of the quantum walk. The convergence rate is now controlled by the minimum spectral gap between \emph{any pair} of distinct eigenvalues of \( H \), not just the smallest nonzero eigenvalue.
\section{Mixing in Classical and Quantum Walks}

\paragraph{Classical Mixing Time}

The \emph{mixing time} of a random walk on a graph quantifies how quickly the probability distribution over vertices converges to the stationary distribution. Consider a continuous-time random walk on a graph \( G = (V, E) \) defined via the Laplacian matrix \( L \). The evolution of the probability vector \( p(t) \in \mathbb{R}^{|V|} \), which represents the probability of being at each vertex at time \( t \), is governed by the differential equation:
\[
\frac{dp(t)}{dt} = Lp(t).
\]
Let \( u := \frac{1}{|V|}(1,1,\dots,1) \) denote the uniform distribution over vertices. This vector is an eigenvector of \( L \) corresponding to eigenvalue zero. Assuming \( G \) is connected, this eigenvalue is unique. Let \( \{ v_\lambda \} \) be an orthonormal basis of eigenvectors of \( L \), where \( \lambda \neq 0 \), such that:
\[
L = \sum_{\lambda \neq 0} \lambda v_\lambda v_\lambda^\top.
\]
Then, the solution to the walk is:
\begin{align}
p(t) &= e^{Lt}p(0) \label{eq:classical1} \\
     &= \left( |V| uu^\top + \sum_{\lambda \neq 0} e^{\lambda t} v_\lambda v_\lambda^\top \right)p(0) \label{eq:classical2} \\
     &= \langle u, p(0) \rangle u + \sum_{\lambda \neq 0} e^{\lambda t} \langle v_\lambda, p(0) \rangle v_\lambda. \label{eq:classical3}
\end{align}
As \( t \to \infty \), each term involving \( \lambda < 0 \) decays exponentially, and \( p(t) \to u \). Hence, the walk asymptotically approaches the uniform distribution. The rate of convergence is governed by the \emph{spectral gap}, i.e., the magnitude of the smallest nonzero eigenvalue of \( L \)~\cite{childs2022quantum}.

\paragraph{Quantum Mixing: Time-Averaged Behavior}

In contrast to classical walks, a quantum walk is a unitary process. As such, it does not converge to a stationary quantum state over time. However, it is still meaningful to define a limiting distribution in the measurement statistics. Consider a quantum walk generated by a Hermitian operator \( H \) (e.g., the Laplacian or adjacency matrix), evolving under:
\[
\ket{\psi(t)} = e^{-iHt} \ket{a},
\]
where \( \ket{a} \) is the initial vertex state. Since unitary evolution is reversible and norm-preserving, convergence is not possible in the traditional sense. Nevertheless, one may define a time-averaged probability of observing vertex \( b \) after evolving for a randomly chosen time \( t \in [0, T] \):
\[
p_{a \to b}(T) = \frac{1}{T} \int_0^T |\langle b | e^{-iHt} | a \rangle|^2 \, dt.
\]
Assuming a non-degenerate spectral decomposition \( H = \sum_\lambda \lambda \ket{\lambda}\bra{\lambda} \), one finds:
\begin{align}
p_{a \to b}(T) &= \sum_\lambda |\langle a | \lambda \rangle|^2 |\langle b | \lambda \rangle|^2 \\
&\quad + \sum_{\lambda \neq \lambda'} \langle b | \lambda \rangle \langle \lambda | a \rangle \langle a | \lambda' \rangle \langle \lambda' | b \rangle \cdot \frac{1 - e^{-i(\lambda - \lambda')T}}{i(\lambda - \lambda')T}.
\end{align}
As \( T \to \infty \), the second term vanishes, and the limiting distribution becomes:
\[
p_{a \to b}(\infty) := \sum_\lambda |\langle a | \lambda \rangle|^2 |\langle b | \lambda \rangle|^2.
\]
This distribution does not depend on time and captures the long-time average behavior of the quantum walk. The convergence rate is now controlled by the minimum spectral gap between \emph{any pair} of distinct eigenvalues of \( H \), not just the smallest nonzero eigenvalue~\cite{childs2022quantum}.

@@

Interestingly, there exist formal \textit{equivalences} between the two models. A coined DTQW can approximate a CTQW through small time-step evolution, as demonstrated by Childs ~\cite{childs2010relationship}, and conversely, CTQWs can often be embedded into DTQWs by carefully engineering the coin space to emulate continuous dynamics. While both are computationally equivalent in power, each model has practical advantages in different settings. CTQWs are frequently preferred for spatial search on low-degree graphs due to the natural formulation of marked Hamiltonians. DTQWs, especially in Szegedy’s coinless formulation, excel in quantizing classical Markov chains and have been instrumental in algorithms such as element distinctness.

@@

@@

@@

Now, to see why we have the above equation, we have:
